The Evolution of Phenotypically Invariant Genetic Networks

### 2018-01-26 kalman_walk_paper.tex
final edits

\begin{abstract}
%The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes.
Even if a species' phenotype does not change over evolutionary time, the underlying mechanism may change, as distinct molecular pathways can realize identical phenotypes.
Here we use quantitative genetics and linear system theory to study
how a gene network underlying a conserved phenotype evolves,
as the genetic drift of small mutational changes to these molecular pathways
cause a population to explore the set of mechanisms with identical phenotypes.
% Treating an organism as a ``black box''
To do this, we model an organism's internal state as a linear system of differential equations
for which the environment provides input and the phenotype is the output,
%which translates some environmental input
%to an output that is subject to the pressures of natural selection-- the phenotype.
in which context there exists 
an exact characterization of the set of all mechanisms that give the same input--output relationship.
%There exists an exact expression for the set of all linear mechanisms with identical phenotypes, 
%Due to the genetic drift of small mutational tweaks to these molecular pathways, we expect an evolving population to explore this set.
This characterization implies that selectively neutral directions in genotype space should be common
% there is never a unique network architecture for any phenotype
and that the evolutionary exploration of these distinct but equivalent mechanisms
can lead to the reproductive incompatibility of independently evolving populations.
This evolutionary exploration, or \emph{system drift}, 
proceeds at a rate proportional to the amount of intrapopulation genetic variation
divided by the effective population size ($N_e$).
At biologically reasonable parameter values
this process can lead to substantial interpopulation incompatibility,
and thus speciation, in fewer than $N_e$ generations.
%over a number of generations proportional to the effective population size.
This model also naturally predicts Haldane's rule, 
thus providing another possible explanation
of why heterogametic hybrids tend to be disrupted more often than homogametes during the early stages of speciation.
\end{abstract}

@@

It is well known that in many contexts mathematical models
can fundamentally be \emph{nonidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertainty about an inferred model's parameters or even its claims about
causal structure, despite access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. 
Models with different parameter schemes, or even different mechanics 
can make equally accurate predictions,
but still not actually reflect the internal dynamics of the system being modelled.
In control theory, where electrical circuits and mechanical systems are often the focus, 
it is understood that there can be an infinite number of ``realizations'', 
or ways to reverse engineer the dynamics of a ``black box'',
even if all possible input and output experiments on the ``black box'' are performed 
\citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
The inherent nonidentifiability of chemical reaction networks is sometimes referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.
Finally,
the field of \emph{inverse problems} studies those cases where,
even if a one-to-one mapping between model and behavior is possible in theory,
even tiny amounts of noise can make inference problems nonidentifiable in practice \citep{petrov2005well}.

@@

In this paper we outline a theoretical framework to study the evolution of biological systems, such as gene regulatory networks.
We study the evolution of an optimally adapted population subject to stabilizing selection for phenotype.
Even if the phenotype remains stable over evolutionary time, the underlying mechanism might not remain so,
as many distinct (and mutationally connected) molecular pathways can realize identical phenotypes.
Below, we first apply results from system theory which give
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes.
Since these phenotypically equivalent gene networks are not necessarily compatible with one another, 
system drift may result in reproductive incompatibility between populations 
isolated for a sufficiently long period of time, 
even in the absence of any sort of adaptive, selective, or environmental change. 
We then use quantitative genetic theory to estimate how quickly 
reproductive incompatibility due to system drift will manifest

@@

It is not a new observation that there is often more than one way to do the same thing, 
or that speciation can be the result of (nearly) neutral processes.
This has been analyzed for traits under stabilizing selection determined additively by alleles at many loci 
\citep{wright1935evolution,barton1986maintenance,barton2001role}
and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
\citet{phillips1996maintenance} showed that population structure
can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes.
Recent work has shown that Fisher's geometric model of stabilizing selection
predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
We use similar tools in the second half of this paper,
differing in that we have a submanifold of optima rather than a single point.
Using simulations, \citet{tulchinsky2014hybrid}
%(and similarly \citet{khatri2015simple})
looked for hybrid incompatibility
following regulatory sequence drift,
and \citet{porter2002speciation} studied speciation in populations expressing a simple regulatory cascade.
Both found rapid speciation following directional selection,
but only equivocal support for speciation under models of purely neutral drift.
%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
Our model differs from previous work in that we use linear systems theory
to both explore a much richer class of regulatory network models,
and provide analytical expectations in large populations with complex phenotypes
that would be inaccessible to population simulations.

@@

Although many different biological systems can be modeled with this approach, for clarity, we focus on gene regulatory networks.
In terms of gene regulatory networks, $A_{ij}$ determines how the $j^\text{th}$ transcription factor regulates the $i^\text{th}$ transcription factor.
If $A_{ij} > 0$, then $\kappa_j$ upregulates $\kappa_i$, while if $A_{ij} < 0$, then $\kappa_j$ downregulates $\kappa_i$.
The $i^\text{th}$ row of $A$ is therefore determined by genetic features such as
the strength of $j$-binding sites in the promoter of gene $i$,
factors affecting chromatin accessibility near gene $i$,
or basal transcription machinery activity.
The form of $B$ determines how the environment influences transcription factor expression levels,
and $C$ might determine the rate of production of downstream enzymes.

%% {Linearity} also ``zero-state equivalence''
Here we have assumed that the system is linear,
and begins from the ``zero'' state ($\kappa(0)=0$).
Of course, neither of these are necessarily true for real systems,
but the dynamics of most nonlinear systems can be approximated locally by a linear systems near most points. 
Furthermore, the ease of analyzing linear systems makes this an attractive place to start.
To demonstrate this approach, we apply it to construct a simple gene network in Example \ref{ex:oscillator} below.

@@

It turns out that in general, there are more degrees of freedom,
except if the system is \emph{minimal} -- meaning, informally, that it uses the smallest possible number of components
to achieve the desired dynamics.
Results in system theory show that any system can be realized in a particular minimal dimension
(the dimension of the kryptotype, $n_\text{min}$),
and that any two phenotypically equivalent systems of dimension $n_\text{min}$ are related by a change of coordinates.
Since gene networks can grow or shrink following gene duplications and deletions, 
these additional degrees of freedom can apply in principle to any system.

@@

Exactly how the coefficients 
(i.e., entries of $A$, $B$ or $C$)
of a haploid system inherited by an offspring from her diploid parent
are determined by the parent's two systems
depends on the genetic basis of any variation in the coefficients.
Thanks to the randomness of meiotic segregation,
the result is random to the extent that each parent is heterozygous
for alleles that affect the coefficients.
Since the $i^\text{th}$ row of $A$ summarizes how each gene regulates gene $i$,
and hence is determined by the promoter region of gene $i$,
the elements of a row of $A$ to tend to be inherited together,
which will create covariance between entries of the same row.
It is, however, a quite general observation that the variation seen among recombinant systems
is proportional to the difference between the two parental systems.
This is certainly true if each coefficient is determined by a single nonrecombining locus,
so that each coefficient in the system produced by meiosis is an independent random choice
between the two parental coefficients.
On the other hand, if a coefficient is determined by additive contributions of independently segregating loci,
then the law of total variance applied after conditioning on the parental origin of each allele
implies that the variance is equal to one-quarter of the squared difference between parental coefficients.

@@

  \begin{figure}[H]
    \centering
    %\begin{tabular}{cc}
    %\includegraphics[width=0.4\textwidth]{figures/f1_quartic2} &
    %\includegraphics[width=0.4\textwidth]{figures/f2_quad2}
    %\end{tabular}
    \includegraphics[width=0.5\textwidth]{examples/F2_vs_F1_divergence_tau0}
   % \caption{
    %\textbf{(left)} 
    \caption{Mean hybrid phenotypic distance from optimum computed with equation \ref{eqn:distance}, using $\rho(t) = \exp(-t/4\pi)$
    for $F_1$ (black) and $F_2$ (blue) hybrids between $A(0)$ and $A(\epsilon)$ parent oscillators. Genetic distance is computed as 
    %Frobenius norm
    %Euclidean distance
    $\sqrt{\sum_{ij} (A_{ij}(0) - A_{ij}(\epsilon))^2}$
    %between two systems
    .}
    %\textbf{(right)} Same, but for $F_2$ offspring (one line per type of $F_2$).
    %\plri{make this $D$ not $D^2$ and say which line is the overall average $F_2$.
    %}
    \label{fig:osc_incompat}
  \end{figure}
\end{example}

\paragraph{Haldane's rule}
This model naturally predicts
Haldane's rule, the observation that if only one hybrid sex is sterile or inviable 
it is likely the heterogametic sex (e.g., the male in XY sex determination systems) \citep{haldane1922sex, orr1997haldane}.
For example, consider an XY species with a two-gene network where the first gene resides on an autosome and the second gene on the X chromosome.
A male whose pair of haplotypes is 
$\left( \left[ \begin{smallmatrix} A_1 & A_2 \\ \cdot & \cdot \end{smallmatrix} \right], \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right] \right)$
has phenotype determined by $A = \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right]$, 
if dosage compensation upregulates heterogametes by a factor of 2 relative to homogametes,
% as in \emph{Drosophila}
while a female homozygous for the haplotype
$\left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$, has phenotype determined by
$A = \left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
An $F_1$ male offspring of these two will have its phenotype determined by
$\left[\begin{smallmatrix} (A_1 + \bar A_1)/2 & (A_2 + \bar A_2)/2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
If both genes resided on the autosomes this system would only be possible in an $F_2$ cross.
More generally, 
if the regulatory coefficients for a system are shared between the sex and one or more autosomal chromosomes, 
$F_1$ males are effectively equivalent to purely autosomal-system $F_2$ hybrids. 
Although many alleles will be dominant if the phenotype--fitness relationship is convex,
the underlying mechanism does not depend on the \emph{dominance theory} \citet{turelli1995dominance} to explain Haldane's rule:
instead, it derives from the nature of segregation load.
% In this setting, Haldane's rule is mechanistically similar to \emph{hybrid breakdown}, 
% the observation that $F_2$ hybrids will often be less fit than $F_1$ hybrid crosses, in the early stages of speciation.
% This mechanism relies on the genetic variation of a system being shared across the sex chromosomes and the autosomes,
% and appears to be consistent with the ``dominance'' theory \citep{orr1997haldane, coyne1998evolutionary, turelli1995dominance} which is often invoked to explain Haldane's rule.
% \plr{Explain that the effects on fitness are dominant.}
%and makes several different predictions than the dominance theory as formulated by \citet{turelli1995dominance}.
%\plr{what are they?}
%\citet{turelli1995dominance} employ a population genetic model which strictly requires alleles causing hybrid incompatibility on the X chromosome not to exhibit additive effects, that is the \emph{dominance coefficient} (typically $h$ in classic population genetics) must not be equal to $\frac{1}{2}$. In their model, if alleles are strictly additive, that is neither dominant nor recessive, then male and female $F_1$ hybrids will drop in fitness at the exact same rates, and therefore defy Haldane's rule.

@@

Thus far we have seen that many distinct molecular mechanisms can realize identical phenotypes
and that these mechanisms may fail to produce viable hybrids.
This leads to the question: does evolution shift molecular mechanisms
fast enough to be a significant driver of speciation?
To approach this question,
we explore a general quantitative genetic model in which a population drifts stochastically
near a set of equivalent and optimal systems
due to the action of recombination, mutation, and demographic noise.
Although this is motivated by the results on linear systems above,
the quantitative genetics calculations are more general,
and only depend on the presence of genetic variation and a continuous set of phenotypically equivalent systems.

@@

Next, we calculate with some simplifying assumptions to give the general idea;
multivariate derivations appear in Appendix \ref{ss:quant_gen}.
There will in general be different amounts of variation in different directions;
to keep the discussion intuitive, we only discuss $\sigma_N$, the amount of variation in ``neutral'' directions
(i.e., directions along $\allS$),
and $\sigma_S$, the amount of variation in ``selected'' directions (perpendicular to $\allS$).
The other relevant scale we denote by $\gamma$,
which the scale on which distance to phenotypic optimum changes as $x$ moves away from the optimal set, $\allS$.
Concretely, $\gamma$ is
%the inverse of the derivative of 
$1/(\frac{d}{du}D(x+uz))$ 
where $x$ is optimal and $z$ is a ``selected'' direction perpendicular to $\allS$.
With these parameters, a typical individual will have a fitness of around $\exp(-(\sigma_S/\gamma)^2)$.
Of course, there are in general many possible neutral and selected directions;
we take $\gamma$ to be an appropriate average over possible directions.

@@

\begin{figure}[H]
\begin{center}
\includegraphics{speciation_rates}
\caption{
Mean drop of $F_1$ and $F_2$ fitness relative to parental species,
with $\omega=1$ and 
\textbf{(A)} $\sigma_N^2 = \sigma_S^2 = \gamma^2$
\textbf{(B)} $\sigma_N^2 = \sigma_S^2 = 0.1 \gamma^2$
\textbf{(C)} $0.1 \sigma_N^2 = \sigma_S^2 = \gamma^2$.
The $F_2$ fitness is from equation \eqref{eqn:fit2},
and the $F_1$ fitness is determined only by the exponential term of that equation.
is small relative to that of increased variance.
\label{fig:speciation_rates}}
\end{center}
\end{figure}

@@

Variation in binding site occupancy may overestimate variation in $A$,
since it does not capture buffering effects (if for instance only one site of many needs to be occupied for transcription to begin),
and variation in expression level measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
Nonetheless, these measures likely give us an idea of the scale of variability.
\citet{kasowski2010variation} found differential occupancy in 7.5\%
of binding sites of a transcription factor (p65) between human individuals.
\citet{verlaan2009targeted} showed that cis-regulatory variation
accounts for around 2--6\% of expression variation in human blood-derived primary cells,
while \citet{lappalainen2013transcriptome} found that human population variation 
explained about 3\% of expression variation.
Allele-specific expression is indicative of standing genetic \emph{cis}-regulatory variation;
\citet{wang2017bayesian} found allele-specific expression in 7.2--8.5\% of transcripts of a flycatcher species,
while \citet{tung2015genetic} observed allele-specific expression in 23.4\% of genes studied in a baboon species. 
Taken together, this suggests that variation in the entries of $A$
may be on the order of at least a few percent between individuals of a population --
doubtless varying substantially between species and between genes.

@@

In this paper, we have used tools from quantitative genetics and control theory
to study the evolution of a mechanistic model of the molecular genotype-phenotype map under stabilizing selection,
providing an explicit model of
phenogenetic drift \citep{weiss2000phenogenetic} and developmental system drift \citep{true2001developmental}.
In this context, the Kalman decomposition \citep{kalman1963mathematical}
gives an analytical description of all phenotypically equivalent gene networks.
It also implies that nearly all systems are nonidentifiable,
and hence the general existence of axes of genetic variation that are not constrained by selection.
The independent movement of separated populations along these axes by genetic drift
can lead to reduction in hybrid viability, and hence speciation,
at a speed that depends on effective population size and amount of genetic variation. 
In this model, at biologically reasonable parameter values,
system drift is a significant -- and possibly rapid -- driver of speciation.
This may at first be surprising because
hybrid inviability appears as a consequence of recombining different, yet functionally equivalent, mechanisms.

Consistent with empirical observation of hybrid breakdown \citep[e.g.][]{plotner2017chlorosis}, 
we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s.
% since $F_1$ and $F_2$ phenotypes diverge linearly and at the square root of time relative to parentals, respectively.
%\plr{refer to Turelli here}
% In this model, this is due to recombination shuffling system architectures, leaving the $F_2$ system homozygous for one parent at one particular locus and homozygous for the other parent at another possibly incompatible locus.
Another natural consequence of the model is Haldane's rule,
that if only one $F_1$ hybrid sex is inviable or sterile it is likely to be the heterogametic sex.
This occurs because if the genes underlying a regulatory network are distributed among both autosomes and the sex chromosome, 
then heterogametic $F_1$s show variation similar to that seen in $F_2$ hybrids,
as a result of segregation load \citep[similarly to][]{fraisse2016genetics}.

@@

\section*{Acknowledgements}
We would like to thank Sergey Nuzhdin, Stevan Arnold, Michael Turelli, Patrick Phillips, Erik Lundgren, and Hossein Asgharian for valuable discussion. We would also like to thank Nick Barton, Sarah Signor and Todd Parsons for helpful comments on the manuscript.
%
%\plr{Josh: please fix capitalization in the references.  To do this, anything that needs to be capitalized gets put in curly braces, which tells bibtex to render the capitalization 'as-is'.}

@@

%\section*{Examples}
%\begin{example}[Metabolic network] \label{ex:metabolic}
%Consider an organism that can metabolize two different sugars (present at logarithmic concentrations $u_{1}$ and $u_{2}$), 
%with two enzymes (at log concentrations $\phi_{1}$ and $\phi_{2}$).
%Further suppose that the second sugar is prefered,

>>>>

\begin{abstract}
%The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes.
Even if a species' phenotype does not change over evolutionary time, the underlying mechanism may change, as distinct molecular pathways can realize identical phenotypes.
Here we use quantitative genetics and linear system theory to study
how a gene network underlying a conserved phenotype evolves,
as the genetic drift of small changes to these molecular pathways
cause a population to explore the set of mechanisms with identical phenotypes.
% Treating an organism as a ``black box''
To do this, we model an organism's internal state as a linear system of differential equations
for which the environment provides input and the phenotype is the output,
%which translates some environmental input
%to an output that is subject to the pressures of natural selection-- the phenotype.
in which context there exists 
an exact characterization of the set of all mechanisms that give the same input--output relationship.
%There exists an exact expression for the set of all linear mechanisms with identical phenotypes, 
%Due to the genetic drift of small mutational tweaks to these molecular pathways, we expect an evolving population to explore this set.
This characterization implies that selectively neutral directions in genotype space should be common
% there is never a unique network architecture for any phenotype
and that the evolutionary exploration of these distinct but equivalent mechanisms
can lead to the reproductive incompatibility of independently evolving populations.
This evolutionary exploration, or \emph{system drift}, 
proceeds at a rate proportional to the amount of intrapopulation genetic variation
divided by the effective population size ($N_e$).
At biologically reasonable parameter values
this process can lead to substantial interpopulation incompatibility,
and thus speciation, in fewer than $N_e$ generations.
%over a number of generations proportional to the effective population size.
This model also naturally predicts Haldane's rule, 
thus providing another possible explanation
of why heterogametic hybrids tend to be disrupted more often than homogametes during the early stages of speciation.
\end{abstract}

@@

It is well known that in many contexts mathematical models
can fundamentally be \emph{nonidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertainty about an inferred model's parameters or even its claims about
causal structure, despite access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. 
Models with different parameter schemes, or even different mechanics 
can make equally accurate predictions,
but still not actually reflect the internal dynamics of the system being modelled.
In control theory, where electrical circuits and mechanical systems are often the focus, 
it is understood that there can be an infinite number of ``realizations'', 
or ways to reverse engineer the dynamics of a ``black box'',
even if all possible input and output experiments on the ``black box'' are performed 
\citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
The inherent nonidentifiability of chemical reaction networks is sometimes referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.
Finally,
the field of \emph{inverse problems} studies those cases in which,
despite the existence of a theoretical one-to-one mapping between a model and behavior,
tiny amounts of noise make inference problems nonidentifiable in practice \citep{petrov2005well}.
%even if a one-to-one mapping between model and behavior is possible in theory,
%even tiny amounts of noise can make inference problems nonidentifiable in practice \citep{petrov2005well}.

@@

In this paper we outline a theoretical framework to study the evolution of biological systems, such as gene regulatory networks.
We study the evolution of an optimally adapted population under stable conditions, that is where phenotype is under stabilizing selection.
Even if the phenotype remains stable over evolutionary time, the underlying mechanism might not remain so,
as many distinct (and mutationally connected) molecular pathways can realize identical phenotypes.
Below, we first apply results from system theory which give
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes.
Since these phenotypically equivalent gene networks are not necessarily compatible with one another, 
system drift may result in reproductive incompatibility between sufficiently isolated populations,
%isolated for a sufficiently long period of time, 
even in the absence of any sort of adaptive, selective, or environmental change. 
We then use quantitative genetic theory to estimate how quickly 
reproductive incompatibility due to system drift manifests.

@@

It is not a new observation that there is often more than one way to do the same thing, 
or that speciation can be the result of (nearly) neutral processes.
The potential for speciation has been analyzed in models of 
traits under stabilizing selection determined additively by alleles at many loci 
\citep{wright1935evolution,barton1986maintenance,barton2001role},
in related fitness landscape models \citep{fraisse2016genetics},
and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
% Recent work has shown that Fisher's geometric model of stabilizing selection
% predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
It has also been shown that population structure
can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}.
However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
and a regulatory cascade \citet{porter2002speciation}
found that speciation could be rapid under directional selection
but only equivocal support for speciation under models of purely neutral drift.
%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
Our model differs from previous work in that we use linear systems theory
to both explore a much richer class of regulatory network models,
and provide analytical expectations in large populations with complex phenotypes
that would be inaccessible to population simulations.

@@

% Although many different biological systems can be modeled with this approach, for clarity, we focus on gene regulatory networks.
In terms of gene regulatory networks, $A_{ij}$ determines how the $j^\text{th}$ transcription factor regulates the $i^\text{th}$ transcription factor.
If $A_{ij} > 0$, then $\kappa_j$ upregulates $\kappa_i$, while if $A_{ij} < 0$, then $\kappa_j$ downregulates $\kappa_i$.
The $i^\text{th}$ row of $A$ is therefore determined by genetic features such as
the strength of $j$-binding sites in the promoter of gene $i$,
factors affecting chromatin accessibility near gene $i$,
or basal transcription machinery activity.
The form of $B$ determines how the environment influences transcription factor expression levels,
and $C$ might determine the rate of production of downstream enzymes.
%% {Linearity} also ``zero-state equivalence''
% Here we have assumed that the system is linear,
% and begins from the ``zero'' state ($\kappa(0)=0$).
% Of course, neither of these are necessarily true for real systems,
% but the dynamics of most nonlinear systems can be approximated locally by a linear systems near most points. 
% Furthermore, the ease of analyzing linear systems makes this an attractive place to start.
To demonstrate this approach, we apply it to construct a simple gene network in Example \ref{ex:oscillator} below.

@@

It turns out that in general, there are more degrees of freedom,
except if the system is \emph{minimal} -- meaning, informally, that it uses the smallest possible number of components
to achieve the desired dynamics.
Results in system theory show that any system can be realized in a particular minimal dimension
(the dimension of the kryptotype, $n_\text{min}$),
and that any two phenotypically equivalent systems of dimension $n_\text{min}$ are related by a change of coordinates.
Since gene networks can grow or shrink following gene duplications and deletions, 
these additional degrees of freedom can apply, in principle, to any system.

@@

Exactly how the coefficients 
(i.e., entries of $A$, $B$ and $C$)
of a haploid system inherited by an offspring from her diploid parent
are determined by the parent's two systems
depends on the genetic basis of any variation in the coefficients.
Thanks to the randomness of meiotic segregation,
the result is random to the extent that each parent is heterozygous
for alleles that affect the coefficients.
Since the $i^\text{th}$ row of $A$ summarizes how each gene regulates gene $i$,
and hence is determined by the promoter region of gene $i$,
the elements of a row of $A$ tend to be inherited together,
which will create covariance between entries of the same row.
It is, however, a quite general observation that the variation seen among recombinant systems
is proportional to the difference between the two parental systems.
% This is certainly true if each coefficient is determined by a single nonrecombining locus,
% so that each coefficient in the system produced by meiosis is an independent random choice
% between the two parental coefficients.
% On the other hand, if a coefficient is determined by additive contributions of independently segregating loci,
% then the law of total variance applied after conditioning on the parental origin of each allele
% implies that the variance is equal to one-quarter of the squared difference between parental coefficients.

@@

  \begin{figure}[H]
    \centering
    %\begin{tabular}{cc}
    %\includegraphics[width=0.4\textwidth]{figures/f1_quartic2} &
    %\includegraphics[width=0.4\textwidth]{figures/f2_quad2}
    %\end{tabular}
    \includegraphics[width=0.5\textwidth]{examples/F2_vs_F1_divergence_tau0}
   % \caption{
    %\textbf{(left)} 
    \caption{Mean hybrid phenotypic distance from optimum computed with equation \eqref{eqn:distance}, using $\rho(t) = \exp(-t/4\pi)$
    for $F_1$ (black) and $F_2$ (blue) hybrids between $A(0)$ and $A(\epsilon)$ parent oscillators. Genetic distance is computed as 
    $\left(\sum_{ij} (A_{ij}(0) - A_{ij}(\epsilon))^2\right)^{1/2}$.
    }
    \label{fig:osc_incompat}
  \end{figure}
\end{example}

\paragraph{Haldane's rule}
This model naturally predicts
Haldane's rule, the observation that if only one hybrid sex is sterile or inviable 
it is likely the heterogametic sex (e.g., the male in XY sex determination systems) \citep{haldane1922sex, orr1997haldane}.
For example, consider an XY species with a two-gene network where the first gene resides on an autosome and the second gene on the X chromosome.
A male whose pair of haplotypes is 
$\left( \left[ \begin{smallmatrix} A_1 & A_2 \\ \cdot & \cdot \end{smallmatrix} \right], \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right] \right)$
has phenotype determined by $A = \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right]$, 
  if dosage compensation upregulates heterogametes by a factor of two relative to homogametes (as with \emph{Drosophila}),
while a female homozygous for the haplotype
$\left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$, has phenotype determined by
$A = \left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
An $F_1$ male offspring of these two will have its phenotype determined by
$\left[\begin{smallmatrix} (A_1 + \bar A_1)/2 & (A_2 + \bar A_2)/2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
If both genes resided on the autosomes, this system would only be possible in an $F_2$ cross.
More generally, 
if the regulatory coefficients for a system are shared between the sex and one or more autosomal chromosomes, 
$F_1$ males are effectively equivalent to purely autosomal-system $F_2$ hybrids, and recall that $F_2$s are significantly less fit on average than $F_1$s (see Figure \ref{fig:osc_incompat}).
Although many alleles will be dominant if the phenotype--fitness relationship is convex,
the underlying mechanism does not depend on the \emph{dominance theory} \citet{turelli1995dominance} to explain Haldane's rule:
instead, it derives from the nature of segregation load.
% In this setting, Haldane's rule is mechanistically similar to \emph{hybrid breakdown}, 
% the observation that $F_2$ hybrids will often be less fit than $F_1$ hybrid crosses, in the early stages of speciation.
% This mechanism relies on the genetic variation of a system being shared across the sex chromosomes and the autosomes,
% and appears to be consistent with the ``dominance'' theory \citep{orr1997haldane, coyne1998evolutionary, turelli1995dominance} which is often invoked to explain Haldane's rule.
% \plr{Explain that the effects on fitness are dominant.}
%and makes several different predictions than the dominance theory as formulated by \citet{turelli1995dominance}.
%\plr{what are they?}
%\citet{turelli1995dominance} employ a population genetic model which strictly requires alleles causing hybrid incompatibility on the X chromosome not to exhibit additive effects, that is the \emph{dominance coefficient} (typically $h$ in classic population genetics) must not be equal to $\frac{1}{2}$. In their model, if alleles are strictly additive, that is neither dominant nor recessive, then male and female $F_1$ hybrids will drop in fitness at the exact same rates, and therefore defy Haldane's rule.

@@

Thus far we have seen that many distinct molecular mechanisms can realize identical phenotypes
and that these mechanisms may fail to produce viable hybrids.
Does evolution shift molecular mechanisms
fast enough to be a significant driver of speciation?
To approach this question,
we explore a general quantitative genetic model in which a population drifts stochastically
near a set of equivalent and optimal systems
due to the action of recombination, mutation, and demographic noise.
Although this is motivated by the results on linear systems above,
the quantitative genetics calculations are more general,
and only depend on the presence of genetic variation and a continuous set of phenotypically equivalent systems.

@@

Next, we calculate with some simplifying assumptions to give the general idea;
multivariate derivations appear in Appendix \ref{ss:quant_gen}.
There will in general be different amounts of variation in different directions;
to keep the discussion intuitive, we only discuss $\sigma_N$, the amount of variation in ``neutral'' directions
(i.e., directions along $\allS$),
and $\sigma_S$, the amount of variation in ``selected'' directions (perpendicular to $\allS$).
The other relevant scale we denote by $\gamma$,
which is the scale on which distance to phenotypic optimum changes as $x$ moves away from the optimal set, $\allS$.
Concretely, $\gamma$ is
%the inverse of the derivative of 
$1/(\frac{d}{du}D(x+uz))$ 
where $x$ is optimal and $z$ is a ``selected'' direction perpendicular to $\allS$.
With these parameters, a typical individual will have a fitness of around $\exp(-(\sigma_S/\gamma)^2)$.
Of course, there are in general many possible neutral and selected directions;
we take $\gamma$ to be an appropriate average over possible directions.

@@

\begin{figure}[H]
\begin{center}
\includegraphics{speciation_rates}
\caption{
Mean drop of $F_1$ and $F_2$ fitness relative to parental species,
with $\omega=1$ and 
\textbf{(A)} $\sigma_N^2 = \sigma_S^2 = \gamma^2$
\textbf{(B)} $\sigma_N^2 = \sigma_S^2 = 0.1 \gamma^2$
\textbf{(C)} $0.1 \sigma_N^2 = \sigma_S^2 = \gamma^2$.
The $F_2$ fitness is from equation \eqref{eqn:fit2},
and the $F_1$ fitness is determined only by the exponential term of that equation,
which is small relative to that of increased variance.
\label{fig:speciation_rates}}
\end{center}
\end{figure}

@@

Variation in binding site occupancy may overestimate variation in $A$,
since it does not capture buffering effects (if for instance only one site of many needs to be occupied for transcription to begin),
and variation in expression level measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
Nonetheless, these measures likely give us an idea of the scale of variability.
%\citet{kasowski2010variation} found 
It has been shown that between human individuals, there is differential occupancy in 7.5\%
of binding sites of a transcription factor (p65) \cite{kasowski2010variation}.
%\citet{verlaan2009targeted} showed that 
It has also been inferred that cis-regulatory variation
accounts for around 2--6\% of expression variation in human blood-derived primary cells \cite{verlaan2009targeted}, and
%while \citet{lappalainen2013transcriptome} found 
that human population variation 
explained about 3\% of expression variation \cite{lappalainen2013transcriptome}.
Allele-specific expression is indicative of standing genetic \emph{cis}-regulatory variation;
%\citet{wang2017bayesian} found 
allele-specific expression in 7.2--8.5\% of transcripts of a flycatcher species has been observed \cite{wang2017bayesian}, as well as
%while \citet{tung2015genetic} observed 
allele-specific expression in 23.4\% of genes studied in a baboon species \cite{tung2015genetic}. 
Taken together, this suggests that variation in the entries of $A$
may be on the order of at least a few percent between individuals of a population --
doubtless varying substantially between species and between genes.

@@

In this paper, we use tools from quantitative genetics and control theory
to study the evolution of a mechanistic model of the genotype-phenotype map, 
in which the phenotype is subject to stabilizing selection.
In so doing, we provide an explicit model of
phenogenetic drift \citep{weiss2000phenogenetic} and developmental system drift \citep{true2001developmental}.
In this context, the Kalman decomposition \citep{kalman1963mathematical}
gives an analytical description of all phenotypically equivalent gene networks.
It also implies that nearly all systems are nonidentifiable,
and that, in general, there exist axes of genetic variation unconstrained by natural selection.
The independent movement of separated populations along these axes by genetic drift
can lead to a significant reduction in hybrid viability, and thus precipitate speciation,
at a speed dependent on the effective population size and the amount of genetic variation. 
In this model, at biologically reasonable parameter values,
system drift is a significant -- and possibly rapid -- driver of speciation.
This may be surprising because
hybrid inviability appears as a consequence of recombining different, yet functionally equivalent, mechanisms, and since species are often defined by their unique adaptations or morphologies.

Consistent with empirical observation of hybrid breakdown (e.g., \citet{plotner2017chlorosis}), 
we see that the fitnesses of $F_2$ hybrids drop at a much faster rate than those of $F_1$s.
% since $F_1$ and $F_2$ phenotypes diverge linearly and at the square root of time relative to parentals, respectively.
%\plr{refer to Turelli here}
% In this model, this is due to recombination shuffling system architectures, leaving the $F_2$ system homozygous for one parent at one particular locus and homozygous for the other parent at another possibly incompatible locus.
Another natural consequence of the model is Haldane's rule,
that if only one $F_1$ hybrid sex is inviable or sterile it is likely to be the heterogametic sex.
This occurs because if the genes underlying a regulatory network are distributed among both autosomes and the sex chromosome, 
then heterogametic $F_1$s show variation (and fitnesses) similar to that seen in $F_2$ hybrids.

@@

\section*{Acknowledgements}

    We would like to thank Sergey Nuzhdin, Stevan Arnold, Michael Turelli, Patrick Phillips, Erik Lundgren and Hossein Asgharian for valuable discussion. 
    We would also like to thank Nick Barton, Sarah Signor and Todd Parsons for very helpful comments on the manuscript.
    Work on this project was supported by funds from
    the Sloan Foundation and the NSF (under DBI-1262645) to PR.

@@
